Write-and-f-array: implementation and an application

### 2013-09-01 msc.tex
fixes for comments

\begin{definition}
An implementation of a concurrent object \emph{serializes} iff for every execution of a program with the implementation
substituted for the object:

For every call to object's method that completes, we can define a \emph{serialization point} in its execution
interval, such that all these calls would return the same values if they happened instaneously at their serialization
points.
\end{definition}

@@

Here we differentiate between threads: the index p in \fn{inc} corresponds to the current thread's ID from range
$\{0, \ldots, P-1\}$. Let us first see that this implementation actually serializes to the increment-only counter.
Obviously, we need to choose the call to WRITE as the serialization point of \fn{inc}. From the monotonicity
of entries of T we can infer that \fn{read} returns a value between (inclusive) the number of calls to \fn{inc} that have
serialized before \fn{read} has started and the number of calls serialized before it has finished.
Thus if a call to \fn{read} returns x, there is a point during its
execution when exactly x \fn{inc}s had been serialized. We can choose this point to be the serialization point of
\fn{read}. Note that it need not correspond to an action executed by the implementation of \fn{read}.

@@

For the second extension, let us note that in many cases one doesn't need the \fn{scan} operation to return the whole array, but rather some sort of its digest. The f-arrays\cite{f-array} allow one to do precisely that for digests that can be expressed as
a result of applying an associative operation to all the array's elements (for example, if we need to retrieve the sum of all array's elements). If we call the operator $f$, the following is a specification of an f-array:

@@

We then use this object to implement a fetch-and-add object. Fetch-and-add is a generalization of the increment-only counter we've seen. Its semantics are specified by the listing below:

@@

\begin{lemma}
\label{lem-hist-is-complete}
If a call to \fn{help}() has executed fully (ie. started and finished) after the serialization point of a successful \fn{publish$_\cdot$}$(v, \cdot)$, then $H[v\mod N].v \geq v$.
\end{lemma}
\begin{proof}
Let us first note that $H[v\mod N].v$ is nondecreasing.
It thus suffices to prove that the condition is met at some point before the end of the call to \fn{help}. We will do so by induction on serialization point of the invocation of \fn{publish}.
Consider a call to \fn{publish$_p$}$(v, \cdot)$ that serializes at time $t$ and a call to \fn{help} that starts after $t$. By Lemma~\ref{lem-hist-help-comes} there is a call to \fn{help} that finishes before next serialization point of a successful \fn{publish$_\cdot$} (thus before $S$
is modified) and before $L[p]$ is modified. Without loss of generality we can assume that the call to \fn{help} we are considering satisfies these conditions. By inductive hypothesis and Lemma~\ref{lem-hist-help-comes}, since a successful call to \fn{publish$_\cdot$}$(v-N+1, \cdot)$ has occured before time $t$, $H[v \mod N].first \geq v-N$ at time $t$. Taking into account that $H[i].v \mod N = i$, $H[v \mod N].v = v-N$ or $H[v \mod N] \geq v$ at any later point in time.
Thus CAS in line~\ref{hist-help-cas} either succeeds, which causes the lemma's conclusion to start being satisfied, or fails, which means that the conclusion was already satisfied.
\end{proof} % TODO: time diagram?

\begin{theorem}
The operations in the implementation of the history object serialize to the single-process object with serialization points of \fn{publish$_\cdot$} as posited earlier.
\end{theorem}
\begin{proof}
Note that $S.v$ is equal to the first argument of the latest successful \fn{publish$_\cdot$} at all times.

Let us first consider a call to \fn{get-current} that returns $r$. From Lemma~\ref{lem-hist-is-complete} we know that $r.v \geq v.v$.
By Observation~\ref{obs-hist-is-correct}, \fn{publish$_\cdot$}$(r.v, \cdot)$ happened before the read from $H[v.v\mod N]$. If $r.v = v.v$, then \fn{publish$_\cdot$}$(r.v, \cdot)$ was the most recent successful \fn{publish$_\cdot$} at the time when $S$ was read, so we can serialize the operation there. Otherwise, \fn{publish$_i$}$(r.v, \cdot)$ has happened after that read, so we can serialize the operation just after it.

Let us consider a calls to \fn{publish$_\cdot$}:
\begin{itemize}
\item If a call to \fn{publish$_\cdot$}$(v, \cdot)$ fails in line~\ref{hist-pub-earlier-exit}, then we can see that the most recently serialized (from the point of view of line~\ref{hist-pub-earlier-sp}) successful \fn{publish$_\cdot$} published a version different than $v-1$.
\item If a call to \fn{publish$_\cdot$}$(v, \cdot)$ fails by failing the CAS in line~\ref{hist-pub-later-sp}, then a successful \fn{publish$_\cdot$} has occured after line~\ref{hist-pub-earlier-sp} had been executed, so the most recent successful \fn{publish$_\cdot$} at the time of CAS has published a version later than $v-1$.
\item If a call to \fn{publish$_\cdot$}$(v, \cdot)$ succeeds, then at the time of CAS in line~\ref{hist-pub-later-sp} the most recently published version is $v-1$.
	\end{itemize}
	This suffices to prove that the value returned by \fn{publish$_\cdot$} is always correct wrt. posited
	serialization order. What remains to consider are the calls to \fn{get}:
\begin{itemize}
	\item If a call to \fn{get}$(v)$ fails in line~\ref{hist-get-notyet}, then the most recent successful \fn{publish$_\cdot$} when line~\ref{hist-get-notyet-sp} executed had a smaller version number than requested, so we can serialize the call to \fn{get} at this line.
	\item If a call to \fn{get}$(v)$ fails in line~\ref{hist-get-tooold} because $h.v > v$, then (by Observation~\ref{obs-hist-is-correct}) a successful call to \fn{publish$_\cdot$}$(h.v, \cdot)$ has occured by that time and $h.v \geq v+N$, so if we serialize \fn{get} at that point, it should fail.
	\item If a call to \fn{get}$(v)$ fails in line~\ref{hist-get-tooold} because $h.v < v$, then (by Lemma~\ref{lem-hist-is-complete}) a call to \fn{publish$_\cdot$}$(v, \cdot)$ has not occured before the call to \fn{help} in line~X. We can thus serialize this \fn{get} just before the call to \fn{help} has started.
	\item If a call to \fn{get}$(v)$ succeeds, then by Observation~\ref{obs-hist-is-correct} it returns a value that was published by a successful call to \fn{publish$_\cdot$}$(v, \cdot)$. By Lemma~\ref{lem-hist-is-complete}, version $v+N$ was not published before the call to \fn{help} has started. Thus we can serialize this \fn{get} just after the successfull call to \fn{publish$_\cdot$}$(v, \cdot)$ has occured, or, if it has occured before \fn{get} started, just before the call to \fn{help} started.
\end{itemize}
\end{proof}

Obviously all operations run in $O(1)$ time. Memory complexity is $O(N+P)$, where $N$ and $P$ are the parameters defined at the beginning of this section.
We will now use this object in our main result, the implementation of a \textsc{write-and-f-array}.
\section{The write-and-f-array}

@@

	\Function{write-and-f$_i$}{$T$}
		\State $r \gets f(v[0], v[1],\ldots, v[N-1])$
		\State $v[i] \gets T$
		\State $\text{last\_update}[i] \gets \text{last\_update}[i] + 1$
		\State $\text{last\_version}[i] \gets \fn{version}(\textbf{true})$
		\State $\text{last\_value}[i] \gets r$
		\State \Return $\text{last\_update}[i], \text{last\_version}[i], \text{last\_value}[i]$
	\EndFunction
	\Function{get-last}{$i$}
		\State \Return $\text{last\_update}[i], \text{last\_version}[i], \text{last\_value}[i]$
	\EndFunction
	\Function{read}{}
		\State \Return $\fn{version}(\textbf{false}), f(v[0], v[1],\ldots, v[N-1])$
	\EndFunction
\end{algorithmic}
%\caption{The enriched write-and-f-array specification}
%\label{spec}
%\end{Listing}

One can observe that the version numbers group calls to \fn{write-and-f} into groups of consecutive calls and 
no call to \fn{read} returns a value with only some updates from a group applied.

@@

%\begin{Listing}[!h]
\begin{algorithmic}[1]
	\State\var{S}{\textbf{pair<int v, \T T>}} %TODO: local var name T is misleading. rename?
	\Function{write-and-f$_0$}{T'}
		\State $v, T \gets S.\fn{rd}()$
		\State $S.\fn{wr}((v + 1, T'))$
		\State \Return $v + 1, v + 1, T$
	\EndFunction
	\Function{get-last}{$i$}
		\State $v, T \gets S.\fn{rd}()$
		\State \Return $v, v, T$
	\EndFunction
	\Function{read}{}
		\State \Return $S.\fn{rd}()$
	\EndFunction
\end{algorithmic}
%\caption{The enriched write-and-f-array snapshot implementation for $N=1$}
%\label{impl1}
%\end{Listing}
%\FloatBarrier

@@

\begin{algorithmic}[1]
	\algrestore{impl}
	\Function{help}{x}
		\State $l_c \gets C[side(x)].\fn{get-last}(child\_id(x))$ \label{get-last-recurse}
		\State $v_{current} \gets H.\fn{get-current}().v$
		\State binary search $\{v_{current}-(N+1),\ldots, v_{current}\}$ for first such $v$ that: \label{help-binsch}
		\Statex $H.\fn{get}(v) \neq \none \textbf{ and } H.\fn{get}(v).v[side(x)] \geq l_c.v$ %indent!
		\If{no such $v$ exists}
			\State \Return
		\EndIf
		\State $h_{old} \gets H.\fn{get}(v-1)$ \label{help-getprev}
		\State $h_{new} \gets H.\fn{get}(v)$
		\If{$h_{old} = \none \textbf{ or } h_{new} = \none$}
			\State \Return
		\EndIf
		\If{$side(x) = 0$}
		\State $T \gets f(l_c.T, h_{old}.T[1])$
		\Else
		\State $T \gets f(h_{new}.T[0], l_c.T)$
		\EndIf
		\State $l \gets L[x].\fn{rd}$
		\If{$l.n < l_C.n$} \label{upd-last-alreadydone}
			\State $L[x].\text{CAS}(l, (l_C.n, v, T))$
		\EndIf
	\EndFunction
	\Function{get-last}{x}
		\State $\fn{help}(x)$
		\State \Return $L[x].\fn{rd}$
	\EndFunction
\end{algorithmic}

The use of binary search in line~\ref{help-binsch} warrants explanation. The predicate employed can obviously change its value in time. Thus, the binary search will return $v$ such that the predicate held at one point
for $v$ and didn't at another point for $v-1$. We will see that this is sufficient. Indeed, the result of the binary search will be important only in the cases when the value of the predicate doesn't change during
the search.

@@

\begin{lemma}
\label{lem-versions-increase}
For any two subsequent successful calls to hist.\fn{publish} with values $h_1$ and $h_2$, $h_1.v[i] \leq h_2.v[i]$ for both $i$.
\end{lemma}
\begin{proof}
Assume otherwise. Consider the first pair $p_1, p_2$ of consecutive calls to \fn{publish} that contradicts the Lemma (see Figure~\ref{fig-versions-increase}). The only place \fn{publish} is called is line \ref{upd-publish}
in \fn{update}.
$p_2$ had its most recent call to \fn{get-current} (denoted by $g_2$) in line~\ref{upd-getown} sometime earlier. No successful publish could have occured between $g_2$ and $p_2$; otherwise $p_2$
would have failed.
Thus, $p_1$ must have occured before $g_2$ and so its execution of lines \ref{upd-getch-0} and \ref{upd-getch-1} (denoted by ``read children'' on the diagram) must have occured strictly earlier than
the same for the second call. This together with monotonicity of versions in children delivers the contradiction.
\end{proof}

\begin{figure}
\begin{tikzpicture}
	\atomic{prev_getch}{read children}{1}{1}
	\atomic{prev_pub}{$p_1$}{3}{1}

@@

	\draw[arrows={-triangle 45}] (prev_pub) to (get_ver);
\end{tikzpicture}
\caption{Diagram for proof of Lemma~\ref{lem-versions-increase} ($p_i$ -- call to \fn{publish}; $g_i$ -- call to \fn{get-current} in line~\ref{upd-getown} that corresponds to $p_i$)}
\label{fig-versions-increase}
\end{figure}

\begin{lemma}
	For every execution of \fn{write-and-f$_i$}: during execution of lines \ref{updcall1} and \ref{updcall2} at least one successful call to \fn{update} occurs.
\end{lemma}
\begin{figure}
\begin{tikzpicture}
	\atomic{our_gv1}{$g_1$}{5.5}{0}
	\atomic{bad_pub1}{$p^S_3$}{6.5}{1}
	\atomic{our_pub1}{$p^F_1$}{7.5}{0}
	\draw[arrows={-triangle 45}] (our_gv1) -- (bad_pub1);
	\draw[arrows={-triangle 45}] (bad_pub1) -- (our_pub1);
	\atomic{bad_gv2}{$g_4$}{7}{2}
	\draw[arrows={-triangle 45}] (bad_pub1) -- (bad_gv2);
	\atomic{our_gv2}{$g_2$}{9.5}{0}
	\atomic{bad_pub2}{$p^S_4$}{11}{2}
	\atomic{our_pub2}{$p^F_2$}{12}{0}
	\draw[arrows={-triangle 45}] (our_gv2) -- (bad_pub2);
	\draw[arrows={-triangle 45}] (bad_pub2) -- (our_pub2);

@@

	\draw[dashed] (6,-2) -- (12,-2);
	\draw (6.5,-2) -- (11.5,-2);
\end{tikzpicture}
\caption{Consequences of two failed \fn{update} calls ($g_i$ -- call to \fn{get-current} in line~\ref{upd-getown}, $p^S_{i}$ -- successful call to \fn{publish} in line~\ref{upd-publish}, $p^F_{i}$ -- failed call to \fn{publish} in line~\ref{upd-publish})}
\label{fig-only-two-pubs}
\end{figure}

@@

\begin{lemma}
	If \fn{help} is called after the serialization point of $n^{\text{th}}$ call to \fn{write-and-f$_i$}, then $L[i].n \geq n$ after the call to \fn{help} completes. \label{last-is-complete}
\end{lemma}
\begin{proof}
	We will prove this Lemma by induction on $n$. Let $p(v)$ be the time the successfull call to \fn{publish}$(v, \cdot)$ occured. Consider an \fn{update$_i$} that was serialized at $p(v)$.
	For every $u \equiv i (\mod N)$, a call to $\fn{help}(i)$ starts and finishes between $p(u-1)$ and $p(u)$. Thus, there is such a call that starts and finishes between $p(v)$ and $p(v+N)$.
	Without loss of generality we can assume that the call to \fn{help} from the Lemma's hypothesis executes in this time interval. In that case, the binary search from line~\ref{help-binsch} will not see any
	$\fn{get}(x)$ for $x \geq v-1$ returning \none and will	return version $v$. For the same reason, both \fn{get}s will succeed.

@@

\begin{lemma}
	If a call to $\fn{help}(i)$ sets $L[i]$ to $n, v, T$, then by the time the call has finished: \label{last-is-correct}
\begin{enumerate}
	\item \fn{write-and-f$_i$} has been called at least $n$ times
	\item Let $n_c, v_c, T_c$ be the return value of the $n^{\text{th}}$ call to $C[side(i)].\fn{write-and-f}_{child\_id(p)}$. Let $v'$ and $h_new$ be the version and node value published at the serialization point of the $n^{\text{th}}$
		\fn{write-and-f$_i$} and $h_old$ be the node value published with version $v'-1$. Then $v = v'$ and:
		\begin{itemize}
			\item If $side(p) = 0$ then $T = f(T_c, h_{old}.T[1])$
			\item If $side(p) = 1$ then $T = f(h_{new}.T[0], T_c)$
		\end{itemize}
\end{enumerate}
\end{lemma}
\begin{proof}
	For the first part, we need to note that by the time line~\ref{get-last-recurse} executes, $n^{\text{th}}$ \fn{write-and-f$_{child\_id(i)}$} has already executed in the appropriate child.
	Let us note that if the binary search fails or returns version different from $v'$, \fn{help} will exit early. Indeed, if the binary search returns a different $v$, $\fn{get}(v-1)$ will fail.
	This immediately proves the first part of the Lemma and shows that $h_{old}$ and $h_{new}$ in the code have the same values as $h_{old}$ and $h_{new}$ in the Lemma, which proves the second part.
\end{proof}

@@

	Consider a set of \fn{write-and-f} calls that are serialized together. Let $h_{new}$ and $h_{old}$ be the history elements published, respectively, at the serialization point and most recently before it. Let $a_i$ be the sequence of return values
	of $C[0].\fn{write-and-f}$ calls corresponding to \fn{write-and-f} calls in question, in serialization order. Let $b_i$ be a similar sequence for $C[1]$. By Lemma~\ref{lem-versions-increase} these are exactly the calls to
$C[i].\fn{write-and-f}$ serialized with versions in $\left(h_{old}.v[i], h_{new}.v[i]\right]$. By Lemma~\ref{last-is-correct} the sequence returned
by the \fn{write-and-f} calls is $f(a_0, b_0), $ $f(a_1, b_0),$ $\ldots, $ $f(h_{new}.T[0], b_1), $ $f(h_{new}.T[0], b_2), \ldots$ in the order of serialization. 
Additionally, $f(h_{new}.T[0], h_{new}.T[1])$ is the value that would be returned by any \fn{read} calls until the next serialization point of \fn{write-and-f}.
This suffices to show that the return values of consecutive calls to \fn{write-and-f} are actually the results of applying the updates.
This proves that values returned by \fn{get} and \fn{update} are correct wrt. posited serialization order.

@@

The correctness of the implementation was tested experimentally both on real hardware and by using the Relacy Race Detector\ref{rrd}. Relacy Race Detector is a framework for testing multi-threaded programs in C++11.
It substitues its own
implementation of synchronization primitives and atomic variables and runs user-supplied tests multiple times, with various interleaving of threads.
It can detect data races on non-atomic variables, deadlock conditions, and failed user-supplied invariant and assertion
checks. One notable feature is support for reduced consistency operations -- Relacy can simulate acquire/release semantics, as specified in the C++11 atomic variables library.
Our fetch-and-add implementation can be compiled both to run on bare hardware and to run in Relacy. The use of Relacy not only privided confidence about correctness of the implementation, but also allowed us to downgrade
consistency guarantees of some writes.

>>>>

\begin{definition}
An implementation of a concurrent object \emph{serializes} iff for every execution of a program with the implementation
substituted for the object:
For every call to object's method that completes, we can define a \emph{serialization point} in its execution
interval, such that all these calls would return the same values if they happened instaneously at their serialization
points.
\end{definition}

@@

Here we differentiate between threads: the index $p$ in \fn{inc} corresponds to the current thread's ID from range
$\{0, \ldots, P-1\}$. Let us first see that this implementation actually serializes to the increment-only counter.
Obviously, we need to choose the call to WRITE as the serialization point of \fn{inc}. From the monotonicity
of entries of T we can infer that \fn{read} returns a value between (inclusive) the number of calls to \fn{inc} that have
serialized before \fn{read} has started and the number of calls serialized before it has finished.
Thus if a call to \fn{read} returns $x$, there is a point during its
execution when exactly $x$ \fn{inc}s had been serialized. We can choose this point to be the serialization point of
\fn{read}. Note that it need not correspond to an action executed by the implementation of \fn{read}.

@@

For the second extension, let us note that in many cases one doesn't need the \fn{scan} operation to return the whole array, but rather some sort of its digest. The f-arrays\cite{f-array} allow one to do precisely that for digests expressible as
a result of applying an associative operation to all the array's elements (for example, if we need to retrieve the sum of all array's elements). If we call the operator $f$, the following is a specification of an f-array:

@@

We then use this object to implement a fetch-and-add object. Fetch-and-add is a generalization of the increment-only counter from previous chapters. Its semantics are specified by the listing below:

@@

\begin{lemma}
\label{lem-hist-is-complete}
If a call to \fn{help}() has executed fully (ie. started and finished) after the serialization point of a successful \fn{publish$_\cdot$}$(v, \cdot)$, then $H[v\mod N].v \geq v$.
\end{lemma}
\begin{proof}
Let us first note that $H[v\mod N].v$ is nondecreasing.
It thus suffices to prove that the condition is met at some point before the end of the call to \fn{help}. We will do so by induction on serialization point of the invocation of \fn{publish}.
Consider a call to \fn{publish$_p$}$(v, \cdot)$ that serializes at time $t$ and a call to \fn{help} that starts after $t$. By Lemma~\ref{lem-hist-help-comes} there is a call to \fn{help} that finishes before next serialization point of a successful \fn{publish$_\cdot$} (thus before $S$
is modified) and before $L[p]$ is modified. Without loss of generality we can assume that the call to \fn{help} we are considering satisfies these conditions. By inductive hypothesis and Lemma~\ref{lem-hist-help-comes}, since a successful call to \fn{publish$_\cdot$}$(v-N+1, \cdot)$ has occured before time $t$, $H[v \mod N].first \geq v-N$ at time $t$. Taking into account that $H[i].v \mod N = i$, at any later point in time one of $H[v \mod N].v = v-N$ and $H[v \mod N] \geq v$ will hold.
Thus CAS in line~\ref{hist-help-cas} either succeeds, which causes the lemma's conclusion to start being satisfied, or fails, which means that the conclusion was already satisfied.
\end{proof} % TODO: time diagram?

\begin{theorem}
The operations in the implementation of the history object serialize to the single-process object with serialization points of \fn{publish$_\cdot$} as posited earlier.
\end{theorem}
\begin{proof}
Note that $S.v$ is equal to the first argument of the latest successful \fn{publish} at all times.

Let us first consider a call to \fn{get-current} that returns $r$. From Lemma~\ref{lem-hist-is-complete} we know that $r.v \geq v.v$.
By Observation~\ref{obs-hist-is-correct}, the successful \fn{publish$_\cdot$}$(r.v, \cdot)$ happened before the read from $H[v.v\mod N]$. If $r.v = v.v$, then this call to \fn{publish} was the most recent successful \fn{publish} at the time when $S$ was read, so we can serialize the operation there. Otherwise, it has happened (had its serialization point) after that read, so we can serialize the operation just after it.

Let us consider a calls to \fn{publish$_\cdot$}:
\begin{itemize}
\item If a call to \fn{publish$_\cdot$}$(v, \cdot)$ fails in line~\ref{hist-pub-earlier-exit}, then we can see that the most recently serialized (from the point of view of line~\ref{hist-pub-earlier-sp}) successful \fn{publish} published a version different than $v-1$.
\item If a call to \fn{publish$_\cdot$}$(v, \cdot)$ fails by failing the CAS in line~\ref{hist-pub-later-sp}, then a successful \fn{publish$_\cdot$} has occured after line~\ref{hist-pub-earlier-sp} had been executed, so the most recent successful \fn{publish$_\cdot$} at the time of \fn{CAS} has published a version greater than $v-1$.
\item If a call to \fn{publish$_\cdot$}$(v, \cdot)$ succeeds, then at the time of CAS in line~\ref{hist-pub-later-sp} the most recently published version is $v-1$.
	\end{itemize}
	This suffices to prove that the value returned by \fn{publish} is always correct with respect to posited
	serialization order. What remains to consider are the calls to \fn{get}:
\begin{itemize}
	\item If a call to \fn{get}$(v)$ fails in line~\ref{hist-get-notyet}, then the most recent successful \fn{publish$_\cdot$} when line~\ref{hist-get-notyet-sp} executed had a smaller version number than requested, so we can serialize the call to \fn{get} at this line.
	\item If a call to \fn{get}$(v)$ fails in line~\ref{hist-get-tooold} because $h.v > v$, then (by Observation~\ref{obs-hist-is-correct}) a successful call to \fn{publish$_\cdot$}$(h.v, \cdot)$ has occured by that time and $h.v \geq v+N$, so if we serialize \fn{get} at that point, it should fail.
	\item If a call to \fn{get}$(v)$ fails in line~\ref{hist-get-tooold} because $h.v < v$, then (by Lemma~\ref{lem-hist-is-complete}) a call to \fn{publish$_\cdot$}$(v, \cdot)$ has not occured before the call to \fn{help} in line~X. We can thus serialize this \fn{get} just before the call to \fn{help} has started.
	\item If a call to \fn{get}$(v)$ succeeds, then by Observation~\ref{obs-hist-is-correct} it returns a value that was published by a successful call to \fn{publish$_\cdot$}$(v, \cdot)$. By Lemma~\ref{lem-hist-is-complete}, version $v+N$ was not published before the call to \fn{help} has started. Thus we can serialize this \fn{get} just after the successfull call to \fn{publish$_\cdot$}$(v, \cdot)$ has occured, or, if it has occured before \fn{get} started, just before the call to \fn{help} started.
\end{itemize}
\end{proof}

Obviously all operations run in $O(1)$ time. Memory complexity is $O(N+P)$, where $N$ and $P$ are the parameters defined at the beginning of this section.
We will now use this object in the main result of this work, the implementation of a write-and-$f$-array.
\section{The write-and-f-array}

@@

	\Function{write-and-f$_i$}{$T$}
		\State $v[i] \gets T$
		\State $r \gets f(v[0], v[1],\ldots, v[N-1])$
		\State $\text{last\_update}[i] \gets \text{last\_update}[i] + 1$
		\State $\text{last\_version}[i] \gets \fn{version}(\textbf{true})$
		\State $\text{last\_value}[i] \gets r$
		\State \Return $\text{last\_update}[i], \text{last\_version}[i], \text{last\_value}[i]$
	\EndFunction
	\Function{get-last}{$i$}
		\State \Return $\text{last\_update}[i], \text{last\_version}[i], \text{last\_value}[i]$
	\EndFunction
	\Function{read}{}
		\State \Return $\fn{version}(\textbf{false}), f(v[0], v[1],\ldots, v[N-1])$
	\EndFunction
\end{algorithmic}
%\caption{The enriched write-and-f-array specification}
%\label{spec}
%\end{Listing}

One can observe that the version numbers group calls to \fn{write-and-f} into groups of consecutive calls with no 
intervening calls to \fn{read}.

@@

%\begin{Listing}[!h]
\begin{algorithmic}[1]
	\State\var{S}{\textbf{pair<int v, \T T>}} %TODO: local var name T is misleading. rename?
	\Function{write-and-f$_0$}{T'}
		\State $v, T \gets S.\fn{rd}()$
		\State $S.\fn{wr}((v + 1, T'))$
		\State \Return $v + 1, v + 1, T'$
	\EndFunction
	\Function{get-last}{$i$}
		\State $v, T \gets S.\fn{rd}()$
		\State \Return $v, v, T$
	\EndFunction
	\Function{read}{}
		\State \Return $S.\fn{rd}()$
	\EndFunction
\end{algorithmic}
%\caption{The enriched write-and-f-array snapshot implementation for $N=1$}
%\label{impl1}
%\end{Listing}
%\FloatBarrier

@@

\begin{algorithmic}[1]
	\algrestore{impl}
	\Function{help}{x}
		\State $l_c \gets C[side(x)].\fn{get-last}(child\_id(x))$ \label{get-last-recurse}
		\State $v_{current} \gets H.\fn{get-current}().v$
		\State binary search $\{v_{current}-(N+1),\ldots, v_{current}\}$ for first such $v$ that: \label{help-binsch}
		\Statex $H.\fn{get}(v) \neq \none \textbf{ and } H.\fn{get}(v).v[side(x)] \geq l_c.v$ %indent!
		\If{no such $v$ exists}
			\State \Return
		\EndIf
		\State $h_{old} \gets H.\fn{get}(v-1)$ \label{help-getprev}
		\State $h_{new} \gets H.\fn{get}(v)$
		\If{$h_{old} = \none \textbf{ or } h_{new} = \none$}
			\State \Return
		\EndIf
		\If{$side(x) = 0$}
			\State $T \gets f(l_c.T, h_{old}.T[1])$
		\Else
			\State $T \gets f(h_{new}.T[0], l_c.T)$
		\EndIf
		\State $l \gets L[x].\fn{rd}$
		\If{$l.n < l_C.n$} \label{upd-last-alreadydone}
			\State $L[x].\text{CAS}(l, (l_C.n, v, T))$
		\EndIf
	\EndFunction
	\Function{get-last}{x}
		\State $\fn{help}(x)$
		\State \Return $L[x].\fn{rd}()$
	\EndFunction
\end{algorithmic}

The use of binary search in line~\ref{help-binsch} warrants explanation. The predicate employed can obviously change its value in time. Thus, the binary search will return such $v$ that the predicate held at one point in time
for $v$ and didn't at another point for $v-1$. We will see that this is sufficient. Indeed, the result of the binary search will be important only in the cases when the value of the predicate doesn't change during
the search.

@@

\begin{lemma}
\label{lem-versions-increase}
For any two subsequent successful calls to $H.\fn{publish}$ with values $h_1$ and $h_2$, $h_1.v[i] \leq h_2.v[i]$ for both $i$.
\end{lemma}
\begin{figure}[!h]
\begin{tikzpicture}
	\atomic{prev_getch}{read children}{1}{1}
	\atomic{prev_pub}{$p_1$}{3}{1}

@@

	\draw[arrows={-triangle 45}] (prev_pub) to (get_ver);
\end{tikzpicture}
\caption{Diagram for proof of Lemma~\ref{lem-versions-increase} ($p_i$ -- call to \fn{publish}; $g_i$ -- call to \fn{get-current} in line~\ref{upd-getown} that corresponds to $p_i$).}
\label{fig-versions-increase}
\end{figure}
\begin{proof}
Assume otherwise. Consider the first pair $p_1, p_2$ of consecutive calls to \fn{publish} that contradicts the Lemma (see Figure~\ref{fig-versions-increase}). The only place \fn{publish} is called is line \ref{upd-publish}
in \fn{update}.
$p_2$ had its most recent call to \fn{get-current} (denoted by $g_2$) in line~\ref{upd-getown} sometime earlier. No successful publish could have occured between $g_2$ and $p_2$; otherwise $p_2$
would have failed.
Thus, $p_1$ must have occured before $g_2$ and so its execution of lines \ref{upd-getch-0} and \ref{upd-getch-1} (denoted by ``read children'' on the diagram) must have occured strictly earlier than
the same for the second call. This together with monotonicity of versions in children delivers the contradiction.
\end{proof}

\begin{lemma}
	For every execution of \fn{write-and-f}: during execution of lines \ref{updcall1} and \ref{updcall2} at least one successful call to \fn{update} occurs.
\end{lemma}
\begin{figure}
\begin{tikzpicture}
	\atomic{our_gv1}{$g_1$}{5.5}{0}
	\atomic{bad_pub1}{$p^S_3$}{6.5}{1}
	\atomic{our_pub1}{$p^F_1$}{7.5}{0}
	\draw[arrows={-triangle 45}] (our_gv1) -- (bad_pub1);
	\draw[arrows={-triangle 45}] (bad_pub1) -- (our_pub1);
	\atomic{bad_gv2}{$g_4$}{7}{2}
	\draw[arrows={-triangle 45}] (bad_pub1) -- (bad_gv2);
	\atomic{our_gv2}{$g_2$}{9.5}{0}
	\atomic{bad_pub2}{$p^S_4$}{11}{2}
	\atomic{our_pub2}{$p^F_2$}{12}{0}
	\draw[arrows={-triangle 45}] (our_gv2) -- (bad_pub2);
	\draw[arrows={-triangle 45}] (bad_pub2) -- (our_pub2);

@@

	\draw[dashed] (6,-2) -- (12,-2);
	\draw (6.5,-2) -- (11.5,-2);
\end{tikzpicture}
\caption{Consequences of two failed \fn{update} calls ($g_i$ -- call to \fn{get-current} in line~\ref{upd-getown}, $p^S_{i}$ -- successful call to \fn{publish} in line~\ref{upd-publish}, $p^F_{i}$ -- failed call to \fn{publish} in line~\ref{upd-publish}).}
\label{fig-only-two-pubs}
\end{figure}

@@

\begin{lemma}
	If \fn{help} is called after the serialization point of $n^{\text{th}}$ call to \fn{write-and-f$_i$}, then $L[i].n \geq n$ after the call to \fn{help} completes. \label{last-is-complete}
\end{lemma}
\begin{proof}
	We will prove this Lemma by induction on $n$. Let $p(v)$ be the time the successfull call to \fn{publish}$(v, \cdot)$ occured. Consider an \fn{update$_i$} that was serialized at $p(v)$.
	For every $u \equiv i \left(\mbox{mod } N\right)$, a call to $\fn{help}(i)$ starts and finishes between $p(u-1)$ and $p(u)$. Thus, there is such a call that starts and finishes between $p(v)$ and $p(v+N)$.
	Without loss of generality we can assume that the call to \fn{help} from the Lemma's hypothesis executes in this time interval. In that case, the binary search from line~\ref{help-binsch} will not see any
	$\fn{get}(x)$ for $x \geq v-1$ returning \none and will	return version $v$. For the same reason, both \fn{get}s will succeed.

@@

\begin{lemma}
	If a call to $\fn{help}(i)$ sets $L[i]$ to $n, v, T$, then by the time the call has finished: \label{last-is-correct}
\begin{enumerate}
	\item \fn{write-and-f$_i$} has been called at least $n$ times.
	\item Let $n_c, v_c, T_c$ be the return value of the $n^{\text{th}}$ call to $C[side(i)].\fn{write-and-f}_{child\_id(p)}$. Let $v'$ and $h_{new}$ be the version and node value published at the serialization point of the $n^{\text{th}}$
		\fn{write-and-f$_i$} and $h_{old}$ be the node value published with version $v'-1$. Then $v = v'$ and:
		\begin{itemize}
			\item if $side(p) = 0$ then $T = f(T_c, h_{old}.T[1])$,
			\item if $side(p) = 1$ then $T = f(h_{new}.T[0], T_c)$.
		\end{itemize}
\end{enumerate}
\end{lemma}
\begin{proof}
	For the first part, we need to note that by the time line~\ref{get-last-recurse} executes, $n^{\text{th}}$ \fn{write-and-f$_{child\_id(i)}$} has already executed in the appropriate child.
	Let us note that if the binary search fails or returns version different from $v'$, \fn{help} will exit early. Indeed, if the binary search returns a different $v$, $\fn{get}(v-1)$ will fail.
	This immediately proves the first part of the Lemma and shows that $h_{old}$ and $h_{new}$ in the code have the same values as $h_{old}$ and $h_{new}$ in the Lemma, which proves the second part.
\end{proof}

@@

	Consider a set of \fn{write-and-f} calls that are serialized together. Let $h_{new}$ and $h_{old}$ be the history elements published, respectively, at the serialization point and most recently before it. Let $a_i$ be the sequence of return values
	of $C[0].\fn{write-and-f}$ calls corresponding to \fn{write-and-f} calls in question, in serialization order. Let $b_i$ be a similar sequence for $C[1]$. By Lemma~\ref{lem-versions-increase} these are exactly the calls to
	$C[j].\fn{write-and-f}$ serialized with versions in $\left(h_{old}.v[j], h_{new}.v[j]\right]$ for $j \in \{0, 1\}$. By Lemma~\ref{last-is-correct} the sequence returned
by the \fn{write-and-f} calls is $f(a_0, b_0), $ $f(a_1, b_0),$ $\ldots, $ $f(h_{new}.T[0], b_0), $ $f(h_{new}.T[0], b_1), \ldots$ in the order of serialization. 
Additionally, $f(h_{new}.T[0], h_{new}.T[1])$ is the value that would be returned by any \fn{read} calls until the next serialization point of \fn{write-and-f}.
This suffices to show that the return values of consecutive calls to \fn{write-and-f} are actually the results of applying the updates.
This proves that values returned by \fn{get} and \fn{update} are correct wrt. posited serialization order.

@@

The correctness of the implementation was tested experimentally both on real hardware and by using the Relacy Race Detector\ref{rrd}. Relacy Race Detector is a framework for testing multi-threaded programs in C++11.
It substitues its own
implementation of synchronization primitives and atomic variables and runs user-supplied tests multiple times, with various interleaving of threads.
It can detect data races on non-atomic variables, deadlock conditions, and failed user-supplied invariant and assertion
checks. One notable feature is support for atomic operations with reduced consistency -- Relacy can simulate acquire/release semantics, as specified in the C++11 atomic variables library.
Our fetch-and-add implementation can be compiled both to run on bare hardware and to run in Relacy. The use of Relacy not only provided confidence about correctness of the implementation, but also allowed us to downgrade
consistency guarantees of some writes.
